Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-25 trunk/paper/drinking.tex

Attribute that can be directly linked to a process are the result of
the temporal extension of a process and include process {\em duration}
and {\em temporal location}. A duration is an attribute that
characterizes the temporal extent of a process and is similar to {\em
  Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional entities. A {\em Temporal location} positions the
time interval during which the process occurs with respect to a
reference coordinate system.

However, the majority of attributes that characterize processes are
not based on genuine process attributes alone, but rather relate
attributes of process participants with the duration of a process. In
particular, a {\em rate} typically refers to an attribute of some
entity {\em with respect to an attribute of another entity}, and in
the context of processes, rates often refer to attributes of a process
participant with respect to the duration of the process. For example,
a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

Not all rates of a process depend on attributes of the process
participants. In particular, a {\em frequency of occurrence} or {\em
  event rate} refers to the number of occurrences of a type of process
during the reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process containing the heart
with a duration of one minute). Further attributes that depend on
types of processes with regard to a reference process are {\em
  distribution patterns}, i.e., how the occurrences of processes of a
particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

Related to distribution patterns are {\em changing qualities} of
processes. For example, the rate of heart beating may change ({\em
  increase} or {\em decrease}) throughout the course of the reference
process. A simple analysis of {\em increasing} ({\em decreasing})
rates would be that the rate of a heart beating within the first half
of a process if {\em lower} ({\em higher}) than in the second half of
the process. Mathematically, this process sub-division can be iterated
until we end up with infinitesimally small time intervals.

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em regulation} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

% In line with the GO, we call processes that cause

% [Regulation of X vs. X: regulation modulates the number/amount of
% occurrence...]

We may say that an object has a {\em frequency} or {\em rate} as if
these were attributes of objects at time points. For example, that my
heart {\em now} beats with a frequency of 80 beats per minute, or that
a car is moving with a speed of 280 kilometres per minute {\em at a
  particular point in time} (e.g., as observed with a speed camera)
may be considered attributes of the objects, not attributes of
processes. In our analysis, rates as attributes of objects can be
explicitly defined using rates of processes. For example, the heart
beating frequency of a particular heart $h$ at a time point $t$ is the
frequency of heart beating processes in which $h$ participates at an
infinitesimally small process around $t$ and within a reference
process. The reference process is necessary in order to obtain a value
for a frequency or rate even if the process in question is not
occurring. For example, a rate of cardiac muscle contraction can be
determined even when no cardiac muscle contraction is occurring in
virtue of a reference process with a rate of cardiac muscle
contraction.

The reference process for an object at a time point $t$ is uniquely
determined for continuous processes (the infinitesimally small process
around $t$), and ambiguous for discrete processes.

A similar construction as for continuous processes can be made for
discrete processes. A rate of blood flow is immediately an attribute
of a single heart beat process. However, during a period of time, the
cumulative rate of blood flow can be observed as the sum of the blood
flow rates of individual heart beating processes.

@@

Complex relations between Based on these definition patterns,
interdependencies between process attributes can be observed.

>>>>

Attribute that can be directly linked to a process arise from
processes' temporal extension. For example, a duration is an attribute
that characterizes the temporal extent of a process and is similar to
{\em Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional physical objects. A {\em Temporal location}
positions the time interval at which the process occurs with respect
to a reference coordinate system.

However, the majority of attributes that characterize processes are
not based on these types of process attributes alone, but rather
relate attributes of process participants with the duration of a
process. In particular, a {\em rate} typically refers to an attribute
of some entity {\em with respect to an attribute of another entity},
and in the context of processes, rates often refer to attributes of a
process participant with respect to the duration of the process. For
example, a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

Related to distribution patterns are {\em changing qualities} of
processes. For example, the rate of heart beating may change ({\em
  increase} or {\em decrease}) throughout the course of a reference
process. A simple analysis of {\em increasing} ({\em decreasing})
rates would be that the rate of a heart beating within the first half
of a process is {\em lower} ({\em higher}) than in the second half of
the process. To make such an assertion, we divide a process into two
temporal parts. Mathematically, this process of sub-division can be
iterated until processes occur within infinitesimally small time
intervals.

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em mass flow} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

We may further attribute a {\em frequency} or {\em rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em heart beating} is
actually occurring.  This reference process can be uniquely determined
for continuous processes (such that the frequency or rate of an object
at a time $t$ is the frequency or rate of the infinitesimally small
process around $t$), and is ambiguous for discrete processes (in which
case the reference process must be explicitly stated).

% A similar construction as for continuous processes can be made for
% discrete processes. A rate of blood flow is immediately an attribute
% of a single heart beat process. However, during a period of time, the
% cumulative rate of blood flow can be observed as the sum of the blood
% flow rates of individual heart beating processes.

@@

% Complex relations between Based on these definition patterns,
% interdependencies between process attributes can be observed.
